Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-05-16 manuscript/manuscript.tex
Edited conclusions

We have developed a Python library that allows us to perform forward gravity computations using tesseroids with variable density in depth.
The code numerically approximates the integral that defines the gravity potential, its gradient and the Marussi tensor through the Gauss-Legendre Quadrature (GLQ).
It essentially consists in approximating the gravity fields of the tesseroids with the effect of point masses located on the scaled nodes of the Legendre polynomials.
We state that this method it's well suited to approximate the effect of tesseroids with arbitrary densities, in contrast with the ones that involve Taylor series expansion.

Instead of enhancing the accuracy of the method by increasing the order of the GLQ on each integration, we make use of the preexisting modified adaptive discretization algorithm and a new density-based discretization algodithm.

The former one divides the tesseroid by half if the ratio of the distance to the computation point and its size is lower than a predefined distance-size ratio $D$.
We can summarise that $D$ controls the accuracy of the method: a higher value of $D$ generates more divisions of the tesseroids, thus more point masses, what leads to a more precise approximation.
Nevertheless it is not sufficient to guarantee the accuracy of the method in case of specific density functions as a highly variable exponential or some discontinuous functions.

For this reason we developed a density-based discretization algorithm that divides the tesseroid on the radius at which the \emph{maximum variation} of the density function takes place, only if a certain inequation that relates this \emph{maximum variation} with a predefined delta ratio $\delta$ holds.
Therefore the number of subdivisions is indirectly controlled by this delta ratio $\delta$, and thus the accuracy of the computation.

Because there is no direct relation between the values assigned to $D$ and $\delta$ and the error of the computation, we had to empirically determine the minimum value of $D$ and the maximum value of $\delta$ that produce an acceptable accuracy for each gravity field.

The distance-size ratio determinations have been made through the comparison of the numerical approximations with the analytical solutions for a spherical shell with variable density, which had to be analytically obtained.
These tests have been made without applying the density-based discretization algorithm in order to isolate the former one.
From the wide range of possible density variations, we've chosen a linear, an exponential and a discontinuous density functions.
The comparisons over the first two functions have shown that values of $D$ equal to 1, 2 and 8 for the calculation of the gravity potential, its gradient and the Marussi tensor components, respectively, produce an accuracy of 0.1\% or better.
On the other hand, the algorithm on its own does not guarantee this amount of accuracy in case of highly varying exponential functions or discontinuous densities.

We have performed similar tests for the density-based discretization algorithm comparing the numerical model with the analytical solutions for a spherical shell with various exponential densities: from low to high varying functions.
Each comparison was made for different values of $\delta$, keeping the default values of $D$ previously obtained.
This tests have shown that a value of $\delta = 0.2$ guarantees an accuracy of 0.1\% or better on every case, while the same results can be obtained performing this tests with the discontinuous density function.
The default value $\delta$ is the same for every gravity field computation.

Moreover, we had performed speed comparisons between tesseroids models with homogeneous and variable densities, showing that a high varying density function requires more computation time than a low varying one, as expected.

>>>>

We have developed a new methodology that computes the gravitational generated 
by a tesseroid with variable density in depth.
It numerically solves the integrals that define the gravitational potential, 
its gradient and the Marussi tensor through the Gauss-Legendre Quadrature 
(GLQ), which essentially consists in approximating the gravitational fields of 
the tesseroids with the effect of point masses located on the scaled nodes of 
the Legendre polynomials.

%~ We state that this method it's well suited to approximate the 
%~ effect of tesseroids with arbitrary densities, in contrast with the ones that 
%~ involve Taylor series expansion.

Instead of enhancing the accuracy of the method by increasing the order of the 
GLQ on each integration, we make use of the pre-existing modified adaptive 
discretization algorithm and a new density-based discretization algorithm.

The former one divides the tesseroid by half if the ratio of the distance to 
the computation point and its size is lower than a predefined distance-size 
ratio $D$.
It's intended to minimise the error when the computation point is close to the 
tesseroid.
We can summarise that the distance-size ratio $D$ controls the accuracy of the 
method: a higher value of $D$ generates more divisions of the tesseroids, thus 
more point masses, what leads to a more precise approximation.
Nevertheless, this algorithm is not sufficient to guarantee the accuracy of 
the method in case of tesseroid with variable density.

For this reason we have developed a density-based discretization algorithm that 
divides the tesseroid on the radius at which the \emph{maximum variation} of 
the density function takes place, only if a certain inequation that relates 
this \emph{maximum variation} with a predefined delta ratio $\delta$ holds.
Therefore the number of subdivisions is indirectly controlled by this delta 
ratio $\delta$, and thus the accuracy of the computation.

Because there is no direct relation between the values assigned to $D$ and 
$\delta$ with the error of the computation, we had to empirically determine 
the minimum value of $D$ and the maximum value of $\delta$ that produce an 
acceptable accuracy for each gravitational field, while maintaining an 
efficient computation.
In order to determine default values of $D$ and $\delta$ we have chosen two 
density functions: a linear and an exponential one.

In the linear density case, the densty-based discretization algorithm is not 
applied due to how it's build, so only a default value of $D$ had to be 
determined.
By comparing the analytical solutions of the gravitational fields generated by 
a spherical shell with the ones produced by the numerical model for different 
values of $D$, we have obtained default values for the distance-size ratio of 
1, 2 and 8 for the potential, its gradient and the Marussi tensor components, 
respectively.
From these comparisons we can guarantee a precision of 0.1\% or greater on the 
computations of the gravitational fields generated by a tesseroid with a 
linear density.

The exponential density case is more complex, due to the fact that both $D$ 
and $\delta$ must be determined simultaneously.
We have explored point in the $D$, $\delta$ space and computed the percentage 
difference between the analytical solutions for a spherical shell and the 
numerical model.
We had to take into account that the $D$ values cannot be lower than the ones 
determined for the linear density case due to the fact that they must 
guarantee acceptable accuracy for every density function, including the linear 
one.
From these comparisons we have concluded that using the same $D$ values 
obtained in the linear density scenario, a $\delta = 0.2$ is needed for the 
computation of the potential and its gradient components, while a $\delta = 
0.01$ must be used for the Marussi tensor components in order to ensure a 
0.1\% precision.

%~ This distance-size determinations have been carried out by comparing the 
%~ analytical solutions of the gravitational fields generated by a spherical 
%~ shell with a linear density with the gravitational fields generated by the 
%~ numerical model for different values of $D$.
%~ This comparisons were taken into account to obtain a percentage difference 
%~ between the gravitational potential, the $g_z$ component of its gradient and 
%~ the diagonal components of the Marussi tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$).
%~ We have set the default value of density-based discretization as the minimum

%~ The distance-size ratio determinations have been made through the comparison 
%~ of the numerical approximations with the analytical solutions for a spherical 
%~ shell with variable density, which had to be analytically obtained. These 
%~ tests have been made without applying the density-based discretization 
%~ algorithm in order to isolate the former one. From the wide range of possible 
%~ density variations, we've chosen a linear, an exponential and a discontinuous 
%~ density functions. The comparisons over the first two functions have shown 
%~ that values of $D$ equal to 1, 2 and 8 for the calculation of the gravity 
%~ potential, its gradient and the Marussi tensor components, respectively, 
%~ produce an accuracy of 0.1\% or better. On the other hand, the algorithm on 
%~ its own does not guarantee this amount of accuracy in case of highly varying 
%~ exponential functions or discontinuous densities.

%~ We have performed similar tests for the density-based discretization algorithm 
%~ comparing the numerical model with the analytical solutions for a spherical 
%~ shell with various exponential densities: from low to high varying functions. 
%~ Each comparison was made for different values of $\delta$, keeping the default 
%~ values of $D$ previously obtained. This tests have shown that a value of 
%~ $\delta = 0.2$ guarantees an accuracy of 0.1\% or better on every case, while 
%~ the same results can be obtained performing this tests with the discontinuous 
%~ density function. The default value $\delta$ is the same for every gravity 
%~ field computation.

%~ Moreover, we had performed speed comparisons between tesseroids models with 
%~ homogeneous and variable densities, showing that a high varying density 
%~ function requires more computation time than a low varying one, as expected.
